### 2023-03-13 main.tex
dualityupdate

We can write the edge Hamiltonian \eqref{edgeHamiltonian} as a function of the coproduct of the second Casimir of $gl(N)$
\begin{equation}
    C_{2}=\sum_{a,b=1}^{N}E_{ab}E_{ba}
\end{equation}
that acts diagonally as $C_{2}|\mathbf{n}\rangle=\twoj(\twoj+N)|\mathbf{n}\rangle$ and belogs to the center of $gl(N)$ (i.e. it commutes with all the basis elements).  \\
 Considering two sites $x,y\in \mathcal{E}$ the standard coproduct operator is defined as 
\begin{equation}
\begin{split}
\Delta^{x,y}:gl(N)&\to gl(N)\otimes gl(N)\\
E_{ab}&\to E_{ab}^{x}\otimes \mathbbm{1}^{y}+\mathbbm{1}^{x}\otimes E_{ab}^{y}
\end{split}
\end{equation}
therefore, we have 
\begin{equation}
\Delta^{x,y}(C_{2})=\sum_{a,b=1}^{N}\Delta(E_{ab})\Delta(E_{ba})=2\sum_{a,b=1}^{N}E_{ab}^{x}\otimes E_{ba}^{y}+C_{2}^{x}\otimes \mathbbm{1}^{y}+\mathbbm{1}^{x}\otimes C_{2}^{y}
\end{equation}

@@

\section{Duality}
\subsection{Definition of duality}
Consider two Markov processes: $(\eta_{t})_{t\geq 0}$ defined on a state space $\Omega$ and $(\xi_{t})_{t\geq 0}$ defined on a state space $\widetilde{\Omega}$. We say that they are dual, with respect to a duality function $D:\Omega\times \widetilde{\Omega}\to \mathbb{R}$, if $\forall \eta\in\Omega$, $\forall \xi\in\widetilde{\Omega}$ and $\forall t> 0$ we have 
\begin{equation}
    \mathbb{E}_{\eta}\left[D(\eta_{t},\xi)\right]=\mathbb{E}_{\xi}\left[D(\eta,\xi_{t})\right]
\end{equation}
where $\mathbb{E}_{\eta}$ denotes the expectation with respect the law of the Markov process $(\eta_{t})_{t\geq 0}$ initialized with the particle configuration $\eta$, whereas $\mathbb{E}_{\xi}$ denotes the expectation with respect the law of the Markov process $(\xi_{t})_{t\geq 0}$ initialized with the particle configuration $\xi$.
The duality definition can also be formulated as a relation between the generators. Call $\mathcal{L}$ the generator of $(\eta_{t})_{t\geq0}$ and $\widetilde{\mathcal{L}}$ the generator of $(\xi_{t})_{t\geq 0}$, then we say that these two processes are dual with respect the duality function $D:\Omega\times \widetilde{\Omega}\to \mathbb{R}$ if $\forall \eta\in\Omega$ and $\forall \xi\in\widetilde{\Omega}$ if
\begin{equation}\label{dualityRelationGenerator}
    \left(\mathcal{L}D(\cdot,\xi)\right)(\eta)=\left(\widetilde{\mathcal{L}}D(\eta,\cdot)\right)(\xi)
\end{equation}
\textbf{Remark}:
in the specific case where $\mathcal{L}=\widetilde{\mathcal{L}}$ we say that the processes are self-dual with respect to a self-duality function.
\newline
\newline
\textbf{Remark}:
when the state spaces of the dual processes is finite, the generators and the duality function are matrices with elements $\mathcal{L}(\eta,\eta^{'})$, $\widetilde{\mathcal{L}}(\xi,\xi^{'})$ and $D(\eta,\xi)$ for arbitrary $\eta,\eta^{'}\in\Omega$ and $\xi,\xi^{'}\in \widetilde{\Omega}$. Therefore, we can write the duality relation \eqref{dualityRelationGenerator} as 
\begin{equation}
    \sum_{\eta^{'}\in\,\Omega}\mathcal{L}(\eta,\eta^{'})D(\eta^{'},\xi)=\sum_{\xi^{'}\in\, \widetilde{\Omega}}\widetilde{\mathcal{L}}(\xi,\xi^{'})D(\eta,\xi^{'})
\end{equation}
that can be read as
\begin{equation}\label{dualityIntertwines}
    \mathcal{L}D=D\widetilde{\mathcal{L}}^{T}
\end{equation}
where the superscript $T$ denotes the matrix transposition. Therefore, the duality relation \eqref{dualityIntertwines} is an intertwining transformation between two linear operators $\mathcal{L}$ and $\widetilde{\mathcal{L}}$. Working with the Hamiltonian operators the duality relation \eqref{dualityIntertwines} reads as
\begin{equation}\label{DualityRelation}
    H^{T}D=D\widetilde{H}
\end{equation}
where $\widetilde{H}=\widetilde{\mathcal{L}}^{T}$.
\subsection{Duality for the generalized multi-species boundary driven stirring process}
We now state our duality result. The process $|n\rangle$ with Hamiltonian \eqref{} is dual in the sense of \eqref{DualityRelation}, with respect to a process $|\xi\rangle$ defined on the enlarged state space
\begin{equation}
    \widetilde{\Omega}:=\Omega\times \mathbb{N}_{0}^{V}
\end{equation}
with elements
\begin{equation}
    |\xi\rangle=\bigotimes_{x\in V}\left(|r_{1}^{x},\ldots,r_{N}^{x}\rangle\otimes |m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}\rangle\right)
\end{equation}
where $r_{x}^{k}$ denotes the number of dual particles of type $k\in \{1,\ldots,N\}$ at site $x\in V$ and $m_{k}^{\widetilde{x}}$ denotes the number of particles of species $k\in \{2,\ldots,N\}$ at an extra-site $\widetilde{x}$ connected to the site $x\in V$.\\
The Hamiltonian of the dual process reads as
\begin{equation}\label{DualHamiltonian}
    \widetilde{H}=\sum_{x,y\in \mathcal{E}}\omega_{x,y}H_{x,y}+\sum_{x\in V}\Gamma_{x}\widetilde{H}_{x}
\end{equation}

where $H_{x,y}$ is the one defined in \eqref{edgeHamiltonian}, while 
\begin{equation}\label{siteDualHamiltonian}
    \widetilde{H}_{x}=\sum_{i=1}^{N}\alpha_{i}^{x}\sum_{k=2}^{N}\left((a^{\dagger})_{k}^{\widetilde{x}}\,E_{1k}^{x}-E_{kk}^{x}\right)
By introducing the action of a bosonic creation operator $a^{\dagger}|q\rangle=|q+1\rangle$ on a generic vector $|q\rangle$ with $q\in \mathbb{N}_{0}$, we define in \eqref{siteDualHamiltonian} 
$(a^{\dagger})_{k}^{\widetilde{x}}$ as a copy of $a^{\dagger}$ acting on the extra-site $\widetilde{x}$ and on the species $k\in\{2,\ldots,N\}$. \\
The Hamiltonians \eqref{Hamiltonian} and \eqref{DualHamiltonian} are dual with respect to the duality matrix $D$ defined as 
    D=\bigotimes_{x\in V}\left(D_{x}\otimes D_{\widetilde{x}}\right)
We denote by $D_{x}$ the site part. This is defined as 
    D_{x}=R_{x}\exp{(E^{x})}
with $E$
    E^{x}=\sum_{k=2}^{N}E_{k1}^{x}
that implies 
    \exp{(E^{x})}=I+E^{x}
and where $R_{x}$ is the diagonal matrix with elements (up to a finite constant) the inverse of the masses of the reversible measure \eqref{reversibleMeasure}, i.e. 
where, for all $x\in V$, we call $\chi=\{(n_{k}^{x})_{k\in\{1,\ldots,N\}}\in \mathbb{N}_{0}\,:\, \sum_{k=1}^{N}n_{k}^{x}=\nu\}$ the simplex.\\ 
We denote by $D_{\widetilde{x}}$ the extra-site part. It is defined as
\begin{equation}\label{dualityMatrix}
D_{\widetilde{x}}=\sum_{m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}=0}^{\infty}\prod_{k=2}^{N}\left(\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}\right)\langle m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}|
On one hand, the edge part of the dual Hamiltonian \eqref{DualHamiltonian} is a copy of \eqref{edgeHamiltonian}, therefore it performs the stirring dynamic on the graph. On the other hand, the site dual Hamiltonian \eqref{siteDualHamiltonian} replaces a particle of any type $k\in\{2,\ldots,N\}$ at site $x$ with a particle of type $1$ and creates a particle of the same type $k$ at the extra-site $\widetilde{x}$. This means that, eventually, only particles of type $1$ remain in the graph while all other types are absorbed by the extra-sites. Thinking about the type $1$ as the hole, similarly to what happens in the SEP(2j)\cite{GioaGiardinaGibertiFrank}, the graph is eventually emptied. 
\subsection{Proof of the duality relation}\label{subsectionDualityProof}
To prove the duality relation between \eqref{} and \eqref{DualHamiltonian} with the duality matrix \eqref{dualityMatrix} we shall show that 
    H^{T}D=D\widetilde{H}
Since the duality matrix is a tensor product over the sites $x\in V$ it is enough to prove that for the edge Hamiltonians
\begin{equation}\label{edgeDualRealtion}
    H_{x,y}^{T}D=DH_{x,y}\qquad \forall x,y\in \mathcal{E}
and for the site Hamiltonians 
\begin{equation}\label{siteDualRelation}
    H_{x}^{T}D=D\widetilde{H}_{x}\qquad \forall x\in V.
%The key idea behind the proof of \eqref{edgeDualRealtion} is that it is possible to write the Hamiltonian \eqref{} as a function of the second Casimir of the $gl(N)$ Lie algebra. Finding a symmetry $S$ of this element, will imply of having a symmetry of the Hamiltonian. By the reversibility of the process one can make $S$ act on a proper matrix (called cheap duality) and obtain 

@@

The proof of the edge duality relation relies on two facts: first the existence of a diagonal matrix, arising from the reversible measure, that is already a duality for the Hamiltonian (in literature called 'cheap' duality matrix, see \cite{}); second the possibility of finding a proper symmetry of the Hamiltonian.  \\We first introduce the cheap duality matrix 

@@

We now look for a symmetry of the Hamiltonian. As shown in \eqref{}, it is possible to write the Hamiltonian \eqref{edgeHamiltonian} as a function of the coproduct of the second Casimir of $gl(N)$. We say that a matrix $A\in gl(N)$ is symmetry of the Casimir $C_{2}$ if

@@

 $C_{2}$ is a central element of $gl(N)$, then it commutes with every element of the algebra. 

@@

\paragraph{Proof of \eqref{siteDualRelation}} The proof of the site duality relations relies on the Hadamard Lemma that, $\forall E,A \in gl(N)$, reads as 

@@

H_{x}^{T}=\sum_{k,l=0}^{N}\alpha_{k}^{x}\left(E_{kl}^{x}-E_{ll}^{x}\right)^{T}=R_{x}\sum_{k,l=0}^{N}\alpha_{k}^{x}\left(E_{lk}^{x}-E_{ll}^{x}\right)R_{x}^{-1}
we multiply both sides for $R_{x}\exp{(E^{x})}$
    H_{x}^{T}R_{x}\exp{(E^{x})}=R_{x}\sum_{k,\ell=0}^{N}\alpha_{k}^{x}\left(E_{lk}^{x}-E_{ll}^{x}\right)R_{x}^{-1}R_{x}\exp{(E^{x})}=R_{x}\sum_{k,l=0}^{N}\alpha_{k}^{x}\left(E_{lk}^{x}-E_{ll}^{x}\right)\exp{(E^{x})}
we insert the identity $I=\exp{(E^{x})}\exp{(-E^{x})}$ in the right hand side

@@

H_{x}^{T}R_{x}\exp{(E^{x})}=R_{x}\exp{(E^{x})}\exp{(-E^{x})}\sum_{k,l=0}^{N}\alpha_{k}^{x}\left(E_{lk}^{x}-E_{ll}^{x}\right)\exp{(E^{x})}
we apply the Hadamard's formula \eqref{HadamardFormula} to 
    \exp{(-E^{x})}\sum_{k,l=0}^{N}\alpha_{k}^{x}\left(E_{lk}^{x}-E_{ll}^{x}\right)\exp{(E^{x})}
We compute the following quantities by the commutation relations \eqref{}:

@@

    \item for all $k\neq 1$ 
        \exp{(-E^{x})}E_{k1}^{x}\exp{(E^{x})}=E_{k1}^{x},
    \exp{(-E^{x})}E_{kk}^{x}\exp{(E^{x})}=E_{kk}^{x}+E_{k1}^{x}

@@

    \exp{(-E^{x})}E_{1k}^{x}\exp{(E^{x})}=E_{1k}^{x}+E_{11}^{x}-\sum_{a=2}^{N}\left(E_{a1}^{x}+E_{ak}^{x}\right)

@@

\sum_{a=2}^{N}\left[E_{a1}^{x},E_{kk}\right]=\sum_{a=2}^{N}\left(E_{ak}^{x}\delta_{k1}-E_{k1}^{x}\delta_{ak}\right)

@@

\sum_{a=2}^{N}\left[E_{a1}^{x},E_{1k}\right]=\sum_{a=2}^{N}\left(E_{ak}^{x}\delta_{11}-E_{11}^{x}\delta_{ak}\right)=\sum_{a=2}^{N}E_{ak}^{x}-E_{11}^{x};

@@

&\sum_{b=2}^{N}\sum_{a=2}^{N}\left[E_{b1}^{x},E_{ak}^{x}\right]-\sum_{b=2}^{N}\left[E_{b1}^{x},E_{11}^{x}\right]=\sum_{a,b=2}^{N}\left(E_{bk}^{x}\delta_{a1}-E_{a1}^{x}\delta_{bk}\right)-\sum_{b=2}^{N}\left(E_{b1}\delta_{11}-E_{11}\delta_{b1}\right)\\=&-2\sum_{a=2}^{N}E_{a1}^{x};

@@

\item for all $k,l\neq 1$
    \exp{(-E^{x})}E_{lk}^{x}\exp{(E^{x})}=E_{kl}^{x}+E_{l1}^{x}

@@

\sum_{a=2}^{N}\left[E_{a1}^{x},E_{lk}\right]=\sum_{a=2}^{N}\left(E_{ak}^{x}\delta_{l1}-E_{l1}^{x}\delta_{ak}\right)=-E_{l1}^{x};

@@

\sum_{a=2}^{N}\left[E_{a1}^{x},E_{lk}\right]=\sum_{a=2}^{N}\left(E_{ak}^{x}\delta_{11}-E_{l1}^{x}\delta_{a1}\right)=\sum_{a=2}^{N}E_{a1}^{x};

@@

     \sum_{k=1}^{N}\alpha_{k}^{x}\left(E_{1k}+E_{11}^{x}-\sum_{a=2}^{N}(E_{jk}^{x}+E_{l1}^{x})-E_{11}^{x}+\sum_{j=2}^{N}E_{j1}^{x}\right)+\sum_{k=1}^{N}\sum_{l=2}^{N}\left(E_{lk}^{x}+E_{\ell 1}^{x}-E_{\ell\ell}^{x}-E_{l1}^{x}\right)

@@

     \sum_{k=1}^{N}\alpha_{k}^{x}\left(E_{1k}^{x}-\sum_{j=2}^{N}E_{jj}^{x}\right)
Thus, multiplying and dividing by $\sum_{i=1}^{N}\alpha_{i}^{x}$, we rewrite \eqref{intermediateTransposeSite} as

@@

H_{x}^{T}R_{x}\exp{(E^{x})}=R_{x}\exp{(E^{x})}\sum_{i=1}^{N}\alpha_{i}^{x}\sum_{k=1}^{N}\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}\left(E_{1k}^{x}-\sum_{j=2}^{N}E_{jj}\right)
We multiply both sides of \eqref{siteHadamard} by 

@@

D_{x}\sum_{m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}=0}^{\infty}\prod_{k=2}^{N}\left(\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}\right)^{m_{k}^{\widetilde{x}}}\langle m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}|\sum_{i=1}^{N}\alpha_{i}^{x}\sum_{k=1}^{N}\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}\left(E_{1k}^{x}-\sum_{j=2}^{N}E_{jj}\right)
Recalling the action of the bosonic creation operator acting at site $\widetilde{x}$ and on the species $k\in \{2,\ldots\}$ we have that 

@@

&\sum_{m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}=0}^{\infty}\prod_{k=2}^{N}\left(\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}\right)^{m_{k}^{\widetilde{x}}+1}\langle m_{2}^{\widetilde{x}},\ldots,m_{k}^{\widetilde{x}}+1,\ldots,m_{N}^{\widetilde{x}}|(a^{\dagger})_{k}^{\widetilde{x}}E_{1k}^{x}

>>>>

We can write the edge Hamiltonian \eqref{edgeHamiltonian} as a function of the coproduct of the second Casimir of $gl(N)$
\begin{equation}\label{secondCasimir}
    C_{2}=\sum_{a,b=1}^{N}E_{ab}E_{ba}
\end{equation}
that acts diagonally as $C_{2}|\mathbf{n}\rangle=\twoj(\twoj+N)|\mathbf{n}\rangle$ and belogs to the center of $gl(N)$ (i.e. it commutes with all the basis elements).  \\
 Considering two sites $x,y\in \mathcal{E}$ the standard coproduct operator is defined as 
\begin{equation}
\begin{split}
\Delta^{x,y}:gl(N)&\to gl(N)\otimes gl(N)\\
E_{ab}&\to E_{ab}^{x}\otimes \mathbbm{1}^{y}+\mathbbm{1}^{x}\otimes E_{ab}^{y}
\end{split}
\end{equation}
therefore, we have 
\begin{equation}
\Delta^{x,y}(C_{2})=\sum_{a,b=1}^{N}\Delta(E_{ab})\Delta(E_{ba})=2\sum_{a,b=1}^{N}E_{ab}^{x}\otimes E_{ba}^{y}+C_{2}^{x}\otimes \mathbbm{1}^{y}+\mathbbm{1}^{x}\otimes C_{2}^{y}
\end{equation}

@@

\section{Duality}
\subsection{Definition of duality}
Consider two Markov processes: $(\eta_{t})_{t\geq 0}$ defined on a state space $\Omega$ and $(\xi_{t})_{t\geq 0}$ defined on a state space $\widetilde{\Omega}$. We say that they are dual, with respect to a duality function $D:\Omega\times \widetilde{\Omega}\to \mathbb{R}$, if $\forall \eta\in\Omega$, $\forall \xi\in\widetilde{\Omega}$ and $\forall t> 0$ we have 
\begin{equation}
    \mathbb{E}_{\eta}\left[D(\eta_{t},\xi)\right]=\mathbb{E}_{\xi}\left[D(\eta,\xi_{t})\right]
\end{equation}
where $\mathbb{E}_{\eta}$ denotes the expectation with respect the law of the Markov process $(\eta_{t})_{t\geq 0}$ initialized with the particle configuration $\eta$, whereas $\mathbb{E}_{\xi}$ denotes the expectation with respect the law of the Markov process $(\xi_{t})_{t\geq 0}$ initialized with the particle configuration $\xi$.
The duality definition can also be formulated as a relation between the generators. Call $\mathcal{L}$ the generator of $(\eta_{t})_{t\geq0}$ and $\widetilde{\mathcal{L}}$ the generator of $(\xi_{t})_{t\geq 0}$, then we say that these two processes are dual with respect the duality function $D:\Omega\times \widetilde{\Omega}\to \mathbb{R}$ if $\forall \eta\in\Omega$ and $\forall \xi\in\widetilde{\Omega}$ if
\begin{equation}\label{dualityRelationGenerator}
    \left(\mathcal{L}D(\cdot,\xi)\right)(\eta)=\left(\widetilde{\mathcal{L}}D(\eta,\cdot)\right)(\xi)
\end{equation}
\textbf{Remark}:
in the specific case where $\mathcal{L}=\widetilde{\mathcal{L}}$ we say that the processes are self-dual with respect to a self-duality function.
\newline
\newline
\textbf{Remark}:
when the state spaces of the dual processes is finite, the generators and the duality function are matrices with elements $\mathcal{L}(\eta,\eta^{'})$, $\widetilde{\mathcal{L}}(\xi,\xi^{'})$ and $D(\eta,\xi)$ for arbitrary $\eta,\eta^{'}\in\Omega$ and $\xi,\xi^{'}\in \widetilde{\Omega}$. Therefore, we can write the duality relation \eqref{dualityRelationGenerator} as 
\begin{equation}
    \sum_{\eta^{'}\in\,\Omega}\mathcal{L}(\eta,\eta^{'})D(\eta^{'},\xi)=\sum_{\xi^{'}\in\, \widetilde{\Omega}}\widetilde{\mathcal{L}}(\xi,\xi^{'})D(\eta,\xi^{'})
\end{equation}
that can be read as
\begin{equation}\label{dualityIntertwines}
    \mathcal{L}D=D\widetilde{\mathcal{L}}^{T}
\end{equation}
where the superscript $T$ denotes the matrix transposition. Therefore, the duality relation \eqref{dualityIntertwines} is an intertwining transformation between two linear operators $\mathcal{L}$ and $\widetilde{\mathcal{L}}$. Working with the Hamiltonian operators the duality relation \eqref{dualityIntertwines} reads as
\begin{equation}\label{DualityRelation}
    H^{T}D=D\widetilde{H}
\end{equation}
where $\widetilde{H}=\widetilde{\mathcal{L}}^{T}$.
\subsection{Duality for the generalized multi-species boundary driven stirring process}
We now state our duality result. The process $|\mathbf{n}\rangle$ with Hamiltonian \eqref{OriginalHamiltonian} is dual in the sense of \eqref{DualityRelation}, with respect to a process $|\bm{\xi}\rangle$ defined on the enlarged state space
\begin{equation}
    \widetilde{\Omega}:=\Omega\times \mathbb{N}_{0}^{V}
\end{equation}
with elements
\begin{equation}
    |\bm{\xi}\rangle=\bigotimes_{x\in V}\left(|r_{1}^{x},\ldots,r_{N}^{x}\rangle\otimes |m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}\rangle\right)
\end{equation}
where $r_{x}^{k}$ denotes the number of dual particles of type $k\in \{1,\ldots,N\}$ at site $x\in V$ and $m_{k}^{\widetilde{x}}$ denotes the number of particles of species $k\in \{2,\ldots,N\}$ at an extra-site $\widetilde{x}$ connected to the site $x\in V$.\\
The Hamiltonian of the dual process reads as
\begin{equation}\label{DualHamiltonian}
    \widetilde{H}=\sum_{x,y\in \mathcal{E}}\omega_{x,y}H_{x,y}+\sum_{x\in V}\Gamma_{x}\widetilde{H}_{x}
\end{equation}

where $H_{x,y}$ is the one defined in \eqref{edgeHamiltonian}, while 
\begin{equation}\label{siteDualHamiltonian}
    \widetilde{H}_{x}=\sum_{i=1}^{N}\alpha_{i}^{x}\sum_{k=2}^{N}\left((a^{\dagger})_{k}^{\widetilde{x}}\,E_{1k}^{x}-E_{kk}^{x}\right)
\end{equation}
By introducing the action of a bosonic creation operator $a^{\dagger}|q\rangle=|q+1\rangle$ on a generic vector $|q\rangle$ with $q\in \mathbb{N}_{0}$, we define in \eqref{siteDualHamiltonian} 
$(a^{\dagger})_{k}^{\widetilde{x}}$ as a copy of $a^{\dagger}$ acting on the extra-site $\widetilde{x}$ and on the occupation of the species $k\in\{2,\ldots,N\}$. \\
The Hamiltonians \eqref{OriginalHamiltonian} and \eqref{DualHamiltonian} are dual with respect to the duality matrix $D$ defined as 
\begin{equation}
    D=\bigotimes_{x\in V}\left(D_{x}\otimes D_{\widetilde{x}}\right)
\end{equation}
We denote by $D_{x}$ the site part. This is defined as 
\begin{equation}
    D_{x}=R_{x}\exp{(E^{x})}
\end{equation}
with 
\begin{equation}
    E^{x}=\sum_{k=2}^{N}E_{k1}^{x}
\end{equation}
that implies 
\begin{equation}
    \exp{(E^{x})}=I+E^{x}
\end{equation}
and where $R_{x}$ is the diagonal matrix with elements (up to a finite constant) the inverse of the masses of the reversible measure \eqref{reversibleMeasure}, i.e. 
\begin{equation}
    R_{x}=\sum_{n\in\chi}\frac{\prod_{k=1}^{N}n_{k}^{x}}{\nu!}|n_{1}^{x},\ldots,n_{N}^{x}\rangle\langle n_{1}^{x},\ldots,n_{N}^{x}|
\end{equation}
where, for all $x\in V$, we call $\chi=\{(n_{k}^{x})_{k\in\{1,\ldots,N\}}\in \mathbb{N}_{0}\,:\, \sum_{k=1}^{N}n_{k}^{x}=\nu\}$ the simplex.\\ 
We denote by $D_{\widetilde{x}}$ the extra-site part. It is defined as
\begin{equation}\label{dualityMatrix}
D_{\widetilde{x}}=\sum_{m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}=0}^{\infty}\prod_{k=2}^{N}\left(\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}\right)\langle m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}|
\end{equation}
On one hand, the edge part of the dual Hamiltonian \eqref{DualHamiltonian} is a copy of \eqref{edgeHamiltonian}, therefore it performs the stirring dynamic on the graph. On the other hand, the site dual Hamiltonian \eqref{siteDualHamiltonian} replaces a particle of any type $k\in\{2,\ldots,N\}$ at site $x$ with a particle of type $1$ and creates a particle of the same type $k$ at the extra-site $\widetilde{x}$. This means that, eventually, only particles of type $1$ remain in the graph while all other types are absorbed by the extra-sites. Thinking about the type $1$ as the hole  the graph is eventually emptied, similarly to what proved in the SEP(2j)\cite{carinci2013duality}.
\subsection{Proof of the duality relation}\label{subsectionDualityProof}
To prove the duality relation between \eqref{OriginalHamiltonian} and \eqref{DualHamiltonian} with the duality matrix \eqref{dualityMatrix} we shall show that 
\begin{equation}
    H^{T}D=D\widetilde{H}
\end{equation}
Since the duality matrix is a tensor product over the sites $x\in V$ it is enough to prove that for the edge Hamiltonians
\begin{equation}\label{edgeDualRealtion}
    H_{x,y}^{T}D=DH_{x,y}\qquad \forall x,y\in \mathcal{E}
\end{equation}
and for the site Hamiltonians 
\begin{equation}\label{siteDualRelation}
    H_{x}^{T}D=D\widetilde{H}_{x}\qquad \forall x\in V.
\end{equation}
We will prove these two statements seprately in the following.

@@

The proof of the edge duality relation relies on two facts: first the existence of a diagonal matrix, arising from the reversible measure, that is already a duality for the Hamiltonian (in literature called 'cheap' duality matrix, see \cite{giardina2009duality}); second the possibility of finding a proper symmetry of the Hamiltonian.  \\We first introduce the cheap duality matrix 

@@

We now look for a symmetry of the Hamiltonian. As shown in \eqref{hamiltonianCasimir}, it is possible to write the Hamiltonian \eqref{edgeHamiltonian} as a function of the coproduct of the second Casimir\eqref{secondCasimir} of $gl(N)$. We say that a matrix $A\in gl(N)$ is symmetry of $C_{2}$ if

@@

 $C_{2}$ is a central element of $gl(N)$, then it commutes with each element of the algebra. 

@@

\paragraph{Proof of \eqref{siteDualRelation}} The proof of the site duality relations relies on the Hadamard Lemma that reads, $\forall E,A \in gl(N)$, as

@@

H_{x}^{T}=\sum_{k,l=0}^{N}\alpha_{k}^{x}\left(E_{k\ell}^{x}-E_{\ell\ell}^{x}\right)^{T}=R_{x}\sum_{k,l=0}^{N}\alpha_{k}^{x}\left(E_{lk}^{x}-E_{ll}^{x}\right)R_{x}^{-1}
First, we multiply both sides for $R_{x}\exp{(E^{x})}$
    H_{x}^{T}R_{x}\exp{(E^{x})}=R_{x}\sum_{k,\ell=0}^{N}\alpha_{k}^{x}\left(E_{\ell k}^{x}-E_{\ell\ell}^{x}\right)R_{x}^{-1}R_{x}\exp{(E^{x})}=R_{x}\sum_{k,\ell =0}^{N}\alpha_{k}^{x}\left(E_{\ell k}^{x}-E_{\ell\ell}^{x}\right)\exp{(E^{x})}
then,we insert the identity $I=\exp{(E^{x})}\exp{(-E^{x})}$ in the right hand side

@@

H_{x}^{T}R_{x}\exp{(E^{x})}=R_{x}\exp{(E^{x})}\exp{(-E^{x})}\sum_{k,\ell=0}^{N}\alpha_{k}^{x}\left(E_{\ell k}^{x}-E_{\ell\ell}^{x}\right)\exp{(E^{x})}
and finally we apply the Hadamard's formula \eqref{HadamardFormula} to 
    \exp{(-E^{x})}\sum_{k,\ell=0}^{N}\alpha_{k}^{x}\left(E_{\ell k}^{x}-E_{\ell \ell}^{x}\right)\exp{(E^{x})}
We compute the following quantities by the commutation relations \eqref{eq:comgl}:

@@

    \item for all $\ell\neq 1$ 
        \exp{(-E^{x})}E_{\ell1}^{x}\exp{(E^{x})}=E_{\ell1}^{x},
    \exp{(-E^{x})}E_{\ell\ell}^{x}\exp{(E^{x})}=E_{\ell\ell}^{x}+E_{\ell1}^{x}

@@

    \exp{(-E^{x})}E_{1\ell}^{x}\exp{(E^{x})}=E_{1\ell}^{x}+E_{11}^{x}-\sum_{a=2}^{N}\left(E_{a1}^{x}+E_{a\ell}^{x}\right)

@@

\sum_{a=2}^{N}\left[E_{a1}^{x},E_{\ell\ell}\right]=\sum_{a=2}^{N}\left(E_{a\ell}^{x}\delta_{k1}-E_{\ell1}^{x}\delta_{a\ell}\right)

@@

\sum_{a=2}^{N}\left[E_{a1}^{x},E_{1\ell}\right]=\sum_{a=2}^{N}\left(E_{a\ell}^{x}\delta_{11}-E_{11}^{x}\delta_{a\ell}\right)=\sum_{a=2}^{N}E_{a\ell}^{x}-E_{11}^{x};

@@

&\sum_{b=2}^{N}\sum_{a=2}^{N}\left[E_{b1}^{x},E_{a\ell}^{x}\right]-\sum_{b=2}^{N}\left[E_{b1}^{x},E_{11}^{x}\right]=\sum_{a,b=2}^{N}\left(E_{b\ell}^{x}\delta_{a1}-E_{a1}^{x}\delta_{b\ell}\right)-\sum_{b=2}^{N}\left(E_{b1}\delta_{11}-E_{11}\delta_{b1}\right)\\=&-2\sum_{a=2}^{N}E_{a1}^{x};

@@

\item for all $k,\ell\neq 1$
    \exp{(-E^{x})}E_{\ell k}^{x}\exp{(E^{x})}=E_{k\ell}^{x}+E_{\ell 1}^{x}

@@

\sum_{a=2}^{N}\left[E_{a1}^{x},E_{\ell k}\right]=\sum_{a=2}^{N}\left(E_{ak}^{x}\delta_{\ell 1}-E_{\ell 1}^{x}\delta_{ak}\right)=-E_{\ell 1}^{x};

@@

\sum_{a=2}^{N}\left[E_{a1}^{x},E_{\ell k}\right]=\sum_{a=2}^{N}\left(E_{ak}^{x}\delta_{11}-E_{\ell 1}^{x}\delta_{a1}\right)=\sum_{a=2}^{N}E_{a1}^{x};

@@

     \sum_{k=2}^{N}\alpha_{k}^{x}\left(E_{1k}+E_{11}^{x}-\sum_{a=2}^{N}(E_{ak}^{x}+E_{a1}^{x})-E_{11}^{x}+\sum_{j=2}^{N}E_{j1}^{x}\right)+\sum_{k=1}^{N}\sum_{l=2}^{N}\alpha_{k}^{x}\left(E_{lk}^{x}+E_{\ell 1}^{x}-E_{\ell\ell}^{x}-E_{l1}^{x}\right)

@@

\sum_{k=2}^{N}\alpha_{k}^{x}E_{1k}^{x}-\sum_{k=1}^{N}\alpha_{k}^{x}\sum_{\ell=2}^{N}E_{\ell\ell}^{x}
     \sum_{k=2}^{N}\left(\alpha_{k}^{x}E_{1k}^{x}-\sum_{i=1}^{N}\alpha_{i}^{x}E_{kk}^{x}\right)
\sum_{i=1}^{N}\alpha_{i}^{x}\sum_{k=2}^{N}\left(\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}E_{1k}^{x}-E_{kk}^{x}\right)
Thus, we rewrite \eqref{intermediateTransposeSite} as

@@

H_{x}^{T}R_{x}\exp{(E^{x})}=R_{x}\exp{(E^{x})}\sum_{i=1}^{N}\alpha_{i}^{x}\sum_{k=2}^{N}\left(\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}E_{1k}^{x}-E_{kk}^{x}\right)
We multiply both sides of \eqref{siteHadamardI} by 

@@

D_{x}\sum_{m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}=0}^{\infty}\prod_{k=2}^{N}\left(\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}\right)^{m_{k}^{\widetilde{x}}}\langle m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}|\sum_{i=1}^{N}\alpha_{i}^{x}\sum_{k=1}^{N}\left(\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}E_{1k}^{x}-E_{kk}\right)
Recalling the action of the bosonic creation operator acting at site $\widetilde{x}$ and on the species $k\in \{2,\ldots,N\}$ we have that 

@@

&\sum_{m_{2}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}=0}^{\infty}\prod_{k=2}^{N}\left(\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}\right)^{m_{k}^{\widetilde{x}}}\langle m_{2}^{\widetilde{x}},\ldots,m_{k}^{\widetilde{x}},\ldots,m_{N}^{\widetilde{x}}|\frac{\alpha_{k}^{x}}{\sum_{i=1}^{N}\alpha_{i}^{x}}E_{1k}^{x}
